Experiences with Planning for Natural Language Generation

### 2009-05-31 trunk/Papers/ci-spark/introduction.tex
changes to Sections 1 and 2

Natural language generation (NLG; \citealp{reiter00building}) is one
of the major subfields of natural language processing, concerned with
computing natural language sentences or texts that convey a given
piece of information to the user. This task can be viewed as a problem
of achieving a (communicative) goal by successively applying
(communicative) actions, and thus has clear intuitive parallels to
automated planning.

This view of generation as planning has a long tradition in NLG
\citep{perrault80,appelt:planning,hovy88,young94dpocl}. Despite their
theoretical attractiveness, these early approaches never gained
mainstream use in applied systems, in part because of the inefficiency
of the planners that were available at the time. However, the recent
improvements in planner efficiency and expressiveness have sparked a
renewed interest in applying modern planning techniques to NLG---both
on the level of speech act planning
\citep{Steedman-Petrick:07,benotti08b} and and on the level of
sentence generation \citep{KolSto07}. The goal of this paper is to
determine, by looking at some representative generation problems,
whether planning has advanced to the point that this enterprise can be
successful this time---and thus, to determine the usefulness of
current planning technology to an application area that has a long
tradition, but is not currently in the focus of the planning
competitions.

The focus of this paper is twofold. First, we present two generation
problems that have recently been cast as planning problems: the
sentence generation task and the GIVE task. In the sentence generation
task, the goal is to generate a single sentence that expresses a given
meaning. \citet{KolSto07} cast this task as a planning problem, where
the plan encodes the necessary sentence and the actions correspond to
uttering individual words.  In the GIVE domain (``Generating
Instructions in Virtual Environments''), we describe a new shared task
that was recently posed as a challenge for the NLG community
\citep{ByrKolStrCasDalMooObe09}.  GIVE uses planning as part of an NLG
system that generates natural-language instructions to guide a user in
performing a given task in a virtual environment.

Second, we discuss some of our experiences using off-the-shelf
planners in these two domains. In particular, we explore the
efficiency of FF \citep{HoffmannNebel01} and SGPLAN
\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
available, support an expressive subset of the Planning Domain
Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}) for
encoding domains, and have been successful on the benchmarks and
problems in the International Planning Competition (IPC)\footnote{See
  \url{http://ipc.icaps-conference.org/} for information about past
  editions of the IPC.}. We test these planners on a range of problem
instances in our planning domains, and compare our results against an
ad-hoc Java implementation of GraphPlan \citep{Blum1997}.

Overall, our findings are mixed. On the one hand, we demonstrate that
modern planners can easily handle the \emph{search} problems that
arise in NLG on realistic inputs, which is promising given that the
sentence generation task, at least, is NP-complete
\citep{KolStr02}. On the other hand, the same planners spend
tremendous amounts of time on preprocessing: Because our domains use a
large number of actions and constants, FF spends 90\% of its runtime
in the sentence generation domain on grounding out all the actions and
literals. As a consequence, the off-the-shelf planners we looked at
are too slow to be useful in the NLG application -- but this is due to
implementation issues rather than the quality of the search
algorithms. We offer our NLG domains as challenges for the planning
community in the hope that these implementation issues might gain more
attention in the future.

This paper is structured as follows. We begin by briefly reviewing
some of the literature on NLG as planning
(Section~\ref{sec:nlg-as-planning}). Then we define introduce the
planning problems arising in two particular NLG tasks in more detail
in Section~\ref{sec:domains}. We report on some experiments with these
planning problems in Section~\ref{sec:experiments}, and discuss the
results in Section~\ref{sec:discussion}. Section~\ref{sec:conclusion}
concludes.

>>>>

\emph{Natural language generation} (NLG; \citealp{reiter00building}) is one of
the major subfields of natural language processing, concerned with computing
natural language sentences or texts that convey a given piece of information to
an audience. While the output of a generation task can take many forms,
including written text, synthesised speech, or embodied multimodal
presentations, the underlying NLG problem in each case can be modelled as a
problem of achieving a (communicative) goal by successively applying a set of
(communicative) actions. This view of NLG as goal-directed action has clear
parallels to \emph{automated planning}, which seeks to find general techniques
for efficiently solving the action sequencing problem.

Treating generation as planning has a long history in NLG, ranging from the
initial attempts of the field to utilise early planning approaches
\citep{perrault80,appelt:planning,hovy88,young94dpocl}, to the recent surge of
research \citep{Steedman-Petrick:07,KolSto07,benotti08b} seeking to capitalise
on the improvements modern planners offer in terms of efficiency and
expressiveness. This paper attempts to assess the usefulness of current planning
techniques to NLG by investigating some representative generation problems, and
evaluating whether automated planning has advanced to the point that it can
provide solutions to such NLG applications---applications that are not currently
being investigated by mainstream planning research.

The focus of this paper is twofold. First, we present two generation problems
that have recently been cast as planning problems: the sentence generation task
and the GIVE task. In the sentence generation task, we concentrate on generating
a single sentence that expresses a given meaning. In this case, a plan encodes
the necessary sentence with the actions in the plan corresponding to the
utterance of individual words \citep{KolSto07}. In the GIVE domain (``Generating
Instructions in Virtual Environments''), we describe a new shared task that was
recently posed as a challenge for the NLG community
\citep{ByrKolStrCasDalMooObe09}. GIVE uses planning as part of a larger NLG
system for generating natural-language instructions that guide a human user in
performing a given task in a virtual environment.

Second, we discuss our experiences using a set of off-the-shelf planners in
these two generation domains. Among the planners we test, we explore the
efficiency of FF \citep{HoffmannNebel01}---a planner that has arguably had the
greatest impact on recent approaches to deterministic planning---and its many
descendants, such as SGPLAN \citep{hsu06:_new_featur_in_sgplan_for}. All of the
planners we test are freely available, support an expressive subset of the
Planning Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
and have been successful on both standard planning benchmarks and the problems
of the International Planning Competition (IPC).\footnote{See
  \url{http://ipc.icaps-conference.org/} for information about recent and past
  editions of the IPC. A good overview of the deterministic track of the 2004
  instance of the IPC can be found in \citep{Hoffmann-Edelkamp:05}.}
Using these planners---and an ad-hoc Java implementation of GraphPlan
\citep{Blum1997} which serves as a baseline for certain experiments---we perform
a series of tests on a range of problem instances in our NLG domains.

Overall, our findings are mixed. On the one hand, we demonstrate that certain
planners can readily handle the \emph{search} problems that arise in our testing
domains on realistic inputs, which is promising given the challenging nature of
these tasks (e.g., the sentence generation task is NP-complete \citep{KolStr02})
On the other hand, these same planners spend tremendous amounts of time on
\emph{preprocessing} techniques that analyse the problem domain in support of
the search. (For instance, FF spends 90\% of its runtime in the sentence
generation domain on grounding out actions and literals.) As a consequence, the
off-the-shelf planners we investigated are often too slow to be useful in real
NLG applications.  Nevertheless, we offer our NLG domains and experiences as
both challenges and lessons for the planning community, in the hope that the
issues we raise might gain more attention in the future.

The remainder of this paper is structured as follows. In
Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as planning and
briefly review the relevant literature. In Section~\ref{sec:domains}, we
describe a set of planning problems associated with two NLG tasks, namely
sentence planning and instruction generation. In Section~\ref{sec:experiments},
we report on the results of our experiments with these planning problems.
In Section~\ref{sec:discussion} we discuss our results and overall
experiences, and conclude in Section~\ref{sec:conclusion}.

### 2009-05-31 trunk/Papers/ci-spark/nlg-as-planning.tex
changes to Sections 1 and 2

The task of generating natural language from semantic representations
(NLG) is typically split up into two parts: the \emph{discourse
  planning} task, which selects the information that is to be conveyed
and structures it into sentence-sized chunks, and the \emph{sentence
  generation} tasks, which then translates each of these chunks into
natural language sentences. The sentence generation task is often
split up into two parts of its own---the \emph{sentence planning}
task, which enriches the input by e.g.\ determining how to refer to
objects and selects some lexical material, and the \emph{surface
  realization} task, which maps the enriched meaning representation
into a sentence using a grammar. The chain of domain planning,
sentence planning, and surface realization is sometimes called the
``NLG pipeline'' \cite{reiter00building}.

Viewing generation as a planning problem has a long tradition in the
NLG literature.  \citet{perrault80} presented an approach to discourse
planning in which the planning operators represent individual speech
acts such as ``request'' and ``inform''. This idea was later worked
out further, e.g.\ by \citet{young94dpocl}. On the other hand,
researchers such as \citet{appelt:planning} and \citet{hovy88} used
techniques from hierarchical planning to expand a high-level plan
consisting of speech acts into more detailed specifications of
individual sentences. Their systems also covered some aspects of
sentence planning. However, these systems used very expressive logics,
which were designed to reason about beliefs and intentions, to
represent the planning state and the planning operators. Furthermore,
most of them used ad-hoc planning algorithms with rather naive search
strategies, which did not scale to realistic inputs. As a consequence,
the NLG-as-planning approach was mostly marginalized throughout the
1990s.

However, there has recently been a string of publications by various
authors with renewed interest in the generation-as-planning approach,
motivated by the recent development of efficient and expressive
planners. Below (Section~\ref{sec:domain-crisp}) we will look in some
detail at \cite{KolSto07}'s approach to sentence generation (i.e., the
sentence planning and surface realization modules of the pipeline) as
planning. In addition, \citet{Steedman-Petrick:07} revisit the
analysis of indirect speech acts with modern planning technology;
\citet{benotti08b} uses planning to explain the accommodation of
presuppositions; and
\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use
multi-agent planning to model the joint problem-solving behavior of
agents in a situated dialogue. These approaches focus on different
issues than the 1980's NLG-as-planning literature. But the main
difference is that they apply \emph{existing}, well-understood
planning approaches to linguistic problems because these offer new
modeling tools, and because they hope that modern planners can solve
the hard search problems that arise in NLG \citep{KolStr02}
efficiently. The point of this paper is to investigate whether
existing planners fulfill this latter promise.

>>>>

The task of generating natural language from semantic representations (NLG) is
typically split into two parts: the \emph{discourse planning} task, which
selects the information to be conveyed and structures it into sentence-sized
chunks, and the \emph{sentence generation} task, which then translates each of
these chunks into natural language sentences. The sentence generation task is
often divided into two parts of its own---the \emph{sentence planning} task,
which enriches the input by, e.g., determining object references and selecting
some lexical material, and the \emph{surface realization} task, which maps the
enriched meaning representation into a sentence using a grammar. The chain of
domain planning, sentence planning, and surface realization is sometimes called
the ``NLG pipeline'' \citep{reiter00building}.

Viewing generation as a planning problem has a long tradition in the NLG
literature. \citet{perrault80} presented an approach to discourse planning in
which the planning operators represented individual speech acts such as
``request'' and ``inform''. This idea was later expanded, e.g., by
\citet{young94dpocl}. On the other hand, researchers such as
\citet{appelt:planning} and \citet{hovy88} used techniques from hierarchical
planning to expand a high-level plan consisting of speech acts into more
detailed specifications of individual sentences. Their systems also covered some
aspects of sentence planning. However, these systems used very expressive
logics, which were designed to reason about beliefs and intentions, to represent
the planning state and the planning operators. Furthermore, most of them used
ad-hoc planning algorithms with rather naive search strategies, which did not
scale to realistic inputs. As a consequence, the NLG-as-planning approach was
mostly marginalized throughout the 1990s.

More recently, there has been a string of publications by various authors with a
renewed interest in the generation-as-planning approach, motivated by the
ongoing development of increasingly more efficient and expressive planners. For
instance, \cite{KolSto07} propose an approach to sentence generation (i.e., the
sentence planning and surface realization modules of the pipeline) as
planning---an approach we will explore in more detail below
(Section~\ref{sec:domain-crisp}). \citet{Steedman-Petrick:07} revisit the
analysis of indirect speech acts with modern planning technology, viewing the
problem as an instance of the general problem of planning with incomplete
information and sensing actions. In addition, \citet{benotti08b} uses planning
to explain the accommodation of presuppositions, and
\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use multi-agent
planning to model the joint problem-solving behaviour of agents in a situated
dialogue. While these approaches focus on different issues compared to the
1980's NLG-as-planning literature, they all apply \emph{existing},
well-understood planning approaches to linguistic problems, in order to utilise
the rich set of modelling tools provided by modern planners, and in the hope
that such planners can efficiently solve the hard search problems that arise in
NLG \citep{KolStr02}. This paper aims to investigate whether existing planners
achieve this latter goal.
